---
title: Predicting Outcomes in "Best Of Series" - Part 1
author: Nathan Wang-Ly
date: '2021-08-18'
description: Why 3-2 series outcomes are so unlikely.
slug: []
categories: []
tags: []
---

```{r, setup, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
```

Across all types of head-to-head competitions, winners are determined through "Best of" series, where competitors must win the majority of the games in the series. For example, e-sports competitions (e.g., League of Legends, Valorant) commonly use Best of 5 series, where the team that first reaches 3 wins is considered the winner.

Before the start of a series, it is typical for analysts to be asked for their predictions. In the case of a Best of 5 series, analysts can predict three possible series scores: 3-0, 3-1, or 3-2 (assuming they know the stronger team/player). These predictions often strike me as being based on some form of intuition; for example, analysts frequently predict a 3-2 score because two teams seem relatively evenly matched.

This got me thinking - when should you expect a series to go 3-0, 3-1, or 3-2? If you expect a team to have an X% chance of winning each game (or on average, they will win Y out of 10 games), what series score should you predict?

### Calculating the probability of each series outcomes

Let's start by making three assumptions:

1. First, of the two players/teams, you know which team is better.
2. Second, you know (or can estimate) the percentage of games they should win against the other team. For example, Team A should beat Team B 70% of the time.
3. Third (and most critically for these calculations), the likelihood of the team winning each game is independent<sup id="a1">[1](#f1)</sup>.

With these assumptions, we can calculate the probability of each series outcome!

Suppose that you have two teams competing in a Best of 5 series: Team A and Team B. You believe that Team A is the better team, and that their (independent) probability of winning each game is $p$.

#### Outcome 1: 3-0

The first possible series outcome is a 3-0. Expressed in wins (W) and losses (L), this would be a *WWW* series. The probability of Team A winning all 3 games would be $p \times p \times p = p^3$.

#### Outcome 2: 3-1

What about a 3-1 series? Those familiar with binomial theory will recall that the probability of 3 wins and 1 loss would be $p^3 \times (1-p)$. However, we also need to multiply this by the number of ways in which a series could go 3-1.

There happen to be 3 ways that this can happen: *LWWW*, *WLWW*, or *WWLW*. Keep in mind that the series *WWWL* is not possible, as the series would have ended 3-0.

So then, the probability of a 3-1 series is $3p^3(1-p)$.

#### Outcome 3: 3-2

Finally, the series could go 3-2. The probably of 3 wins and 2 losses would be $p^3 \times (1-p)^2$. 

There are 6 ways a series could go 3-2: *LLWWW*, *LWLWW*, *LWWLW*, *WLLWW*, *WLWLW*, *WWLLW*.

Therefore, the probability of a 3-2 series is $6p^3(1-p)^2$.

#### Example

To make these probabilities tangible, let's input in a value for $p$. Let's say that you think Team A has a 75% chance of winning each game. The respective probabilities of the three series outcomes would be:

* 3-0: $0.75^3 = 0.422$, or a 42.2% chance
* 3-1: $3 \times 0.75^3 \times 0.25 = 0.316$, or a 31.6% chance
* 3-2: $6 \times 0.75^3 \times 0.25^2 = 0.158$ or a 15.8% chance

This would mean the most likely series outcome is a 3-0 for Team A! Note that the probabilities here do not sum to 1 (they sum to 0.896) because there is an 11.4% chance that Team A would lose the series, either 0-3, 1-3, or 2-3. 

### Comparing the probabilities of each series outcomes

Now that we know the probability of each outcome as a function of $p$, we can calculate it for all values between 0.5 and 1 (since we assume that the team we have chosen is more likely to win).

This can easily be done in R.

First, create a dataframe with the series scores and probabilities.

```{r}
data <- data.frame(
  # repeat each outcome 51 times
  score = rep(c('3-0', '3-1', '3-2'), each = 51),
  # repeat the sequence of probabilities from 0.5-1 (in 0.01 increments) once for each outcome
  p_win = rep(seq(0.5, 1, 0.01), times = 3)
)

# Preview of first few rows
data %>% 
  head(5)
```

Next, we can apply the probabilities from the earlier section.

```{r}
data <- data %>% 
  mutate(p_outcome = case_when(
    # If 3-0
    score == '3-0' ~ (p_win**3),
    # If 3-1
    score == '3-1' ~ 3*(p_win**3)*(1-p_win),
    # If 3-2
    score == '3-2' ~ 6*(p_win**3)*((1-p_win)**2)
  ))

# Probabilities when p = 0.75
data %>% 
  filter(p_win == 0.75)
```

We can see that the probabilities match what we calculated earlier!

Finally, we can create a plot to help us visualise how these probabilities change.

```{r}
data %>% 
  ggplot(aes(x = p_win, y = p_outcome, colour = score)) + 
  geom_path() + 
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) + 
  labs(x = 'Probability of Team A winning each game',
       y = 'Probability of series outcome',
       colour = 'Series outcome') + 
  theme_bw()
```

### Takeaways

What can we take away from this plot?

When $p$ is 0.50 (you think both teams are exactly evenly matched), it is equally likely that the series will go 3-1 or 3-2. There's an 18.8% chance of each, which beats the 12.5% chance of a 3-0.

```{r}
data %>% 
  filter(p_win == 0.5)
```

If you think the chances of Team A winning each game are between 50% and 66%, the most likely outcome is a 3-1.

* If Team A has a 66% chance of winning, the probability of a 3-1 series (29.3%) is slightly higher than a 3-0 series (28.7%).
* If this increases to 67%, the probability of a 3-0 series (30.1%) overtakes the probability of a 3-1 series (29.8%).

```{r}
data %>% 
  filter(p_win %in% c(0.66, 0.67))
```

If you think Team A has anything higher than a 67% chance of winning, the most likely outcome is a 3-0 series. 

Interestingly, this means, that a 3-2 series is never the single-most likely outcome. Even if you believe that two teams are perfectly matched, it only becomes equally likely as a 3-1 series. (Perhaps this may have been intuitive to others, but it came as a surprise to me!)

### Concluding thoughts

This was an interesting exercise for myself to understand how I should expect a Best of series to go based on my estimates of team/player strengths. It definitely came as a surprise to learn how unlikely 3-2 series are (though this may have been intuitive to others)!

Naturally, the calculations are only specific to a Best of 5 series, and they also assume that your goal is to predict the most likely outcome (which it would be, if you were an analyst).

In the next two parts, we'll expand in two directions:

1. How would we generalise these probability calculations to other (odd) Best of series? E.g., a Best of 3, Best of 7, etc.
2. Would we choose differently if our goal was not to predict the outcome, but maximise our expected value in betting on the right outcome? 

<b id="f1">1</b> In reality, this is unlikely to be the case. Momentum often seems to play a pivotal role in the outcome of a series. For example, if a team loses a long, drawn out game, this may affect their mentality and performance in the subsequent games. [↩](#a1)